Understanding the Lomb-Scargle Periodogram

### 2017-01-04 PracticalLombScargle.tex
Add multiterm figures

\section{Least-Squares Extensions to Lomb-Scargle}
\sectlabel{lomb-scargle-extensions}

\begin{figure}[ht]
  \centering
  \includegraphics[width=\textwidth]{fig18_ls_model}
  \caption{The sinusoidal model implied by the Lomb-Scargle periodogram for
    the LINEAR data seen in \fig{LINEAR-data}.
    Although a sinusoid does not perfectly fit the data, it is close enough
    that it serves to find the correct period without a problem.
    \figlabel{ls-model}}
\end{figure}

The equivalence of the Fourier interpretation and least squares interpretation of the Lomb-Scargle periodogram allows for some interesting and useful extensions to the periodogram.
In the least squares interpretation of the periodogram, a sinusoidal model is proposed at each candidate frequency $f$:
\begin{equation}
  y_{model}(t;f) = A_f \sin(2 \pi f (t - \phi_f))
\end{equation}
where the amplitude $A_f$ and phase $\phi_f$ can vary as a function of frequency.
This is fit to the data in the standard least-squares sense, by constructing
the $\chi^2$ statistic at each frequency:
\begin{equation}
  \chi^2(f) \equiv \sum_n \big(y_n - y_{model}(t_n;f)\big)^2
  \eqlabel{chi2-simple}
\end{equation}
We can find the ``best'' model $\hat{y}(t;f)$ by minimizing $\chi^2(f)$ at
each freuqnecy with respect to $A_f$ and $\phi_f$;
we will denote this minimum value as $\hat{\chi}^2(f)$.
\citet{Scargle82} showed that with this setup, the Lomb-Scargle
periodogram from \eq{lomb-scargle-periodogram} can be equivalently written:
\begin{equation}
  P(f) = \frac{N}{2}\big[\hat{\chi}^2_0 - \hat{\chi}^2(f)\big]
  \eqlabel{lomb-scargle-chi2}
\end{equation}
where $\hat{\chi}^2_0$ is the reference model that assumes no variation in the data.
The key realization here is that the Lomb-Scargle periodogram essentially
{\it assumes a sinusoidal model} for the data; this is visualized in
\fig{ls-model} for the data we had seen in \fig{LINEAR-data}.
This immediately begs the question: can we compute a ``periodogram'' based
on more general forms of $y(t;f)$ to more effectively fit the data?

\subsection{Observational Noise}
\label{extensions-observational-noise}
Perhaps the most immediately obvious modification of the periodogram is to
change not the form of the model $y_{model}$, but the $\chi^2$ statistic itself
to account for observational errors in the data.
If there are Gaussian errors $\sigma_n$ on each observed $y_n$, we can re-write \eq{chi2-simple} in the following (standard) way:
\begin{equation}
  \chi^2(f) \equiv \sum_n \left(\frac{y_n - y_{model}(t_n;f)}{\sigma_n}\right)^2
  \eqlabel{chi2-with-errors}
\end{equation}
The periodogram constructed from this $\chi^2$ definition will more accurately
reflect the spectral power of noisy observations.
The effect of this modification on \eq{lomb-scargle-periodogram} is the addition
of a multiplicative weight $1/\sigma_n$ within each of the summations.
Early versions of this sort of modification appear in \citet{Gilliland87}
and \citet{Irwin89}; \citet{Zechmeister09} showed that such a modification
does not change the desirable statistical properties of the periodogram.

@@

@@

We have noted that the periodogram explored in \citep{Lomb76} and
\citep{Scargle82} is equivalent to computing the $\chi^2$ goodness-of-fit
of a sinusoid at each frequency.
Such a fit requires that the data be pre-centered; that is,
\begin{equation}
  \int_0^P g(t) dt = 0
\end{equation}fitting a sinusoid to the data at each
frrrequires the data to be pre-centered; i.e. the signal
must be modified so that

will only work correctly if the data is pre-centered so
that a sinusoid with no offset will fit the data.
In the past this was primarily accomplished by pre-centering the data using the
mean estimated from the data points themselves.
With full phase coverage of the observed signal, this approach is generally
robust, but due to selection effects and survey cadence, full coverage can
not always be guaranteed.
This can potentially lead to situations where the periodogram entirely misses
peaks of interest; an example of this is shown in \fig{standard-vs-floatingmean}.
This simulates noisy observations of a sinusoidal signal situation in which the faintest parts of the light curve fall below the detection threshold.
Using the standard periodogram, where the mean is estimated from the observations, leads to a periodogram that misses the true period of 0.3 days.
The floating-mean periodogram, on the other hand, fits the mean as part of the model and more accurately characterizes the periodic content of the partially-observed data.

\subsection{Fitting Higher-order Models}

\citet{Bretthorst88}

\todo{Use a multi-band periodogram on Kepler Data}

\todo{Show poor performance on eclipsing data}

@@

>>>>

\section{The Least-Square Periodogram and its Extensions}
\sectlabel{lomb-scargle-extensions}

The equivalence of the Fourier interpretation and least squares interpretation of the Lomb-Scargle periodogram allows for some interesting and useful
extensions, which we will explore in this section.
First, let's look at the Least Squares periodogram itself:

In the least squares interpretation of the periodogram, a sinusoidal model is proposed at each candidate frequency $f$:
\begin{equation}
  y_{model}(t;f) = A_f \sin(2 \pi f (t - \phi_f))
\end{equation}
where the amplitude $A_f$ and phase $\phi_f$ can vary as a function of frequency.
This is fit to the data in the standard least-squares sense, by constructing
the $\chi^2$ statistic at each frequency:
\begin{equation}
  \chi^2(f) \equiv \sum_n \big(y_n - y_{model}(t_n;f)\big)^2
  \eqlabel{chi2-simple}
\end{equation}
We can find the ``best'' model $\hat{y}(t;f)$ by minimizing $\chi^2(f)$ at
each freuqnecy with respect to $A_f$ and $\phi_f$;
we will denote this minimum value as $\hat{\chi}^2(f)$.
\citet{Scargle82} showed that with this setup, the Lomb-Scargle
periodogram from \eq{lomb-scargle-periodogram} can be equivalently written:
\begin{equation}
  P(f) = \frac{N}{2}\big[\hat{\chi}^2_0 - \hat{\chi}^2(f)\big]
  \eqlabel{lomb-scargle-chi2}
\end{equation}
where $\hat{\chi}^2_0$ is the non-varying reference model.
The key realization here is that the Lomb-Scargle periodogram essentially
{\it assumes a sinusoidal model} for the data; this is visualized in
\fig{ls-model} for the data we had seen in \fig{LINEAR-data}.
This immediately begs the question: can we compute a ``periodogram'' based
on more general forms of the above model to more effectively fit the data?

\begin{figure}[ht]
  \centering
  \includegraphics[width=\textwidth]{fig18_ls_model}
  \caption{The sinusoidal model implied by the Lomb-Scargle periodogram for
    the LINEAR data seen in \fig{LINEAR-data}.
    Although a sinusoid does not perfectly fit the data, the sinusoidal model
    is close enough that it serves to locate the correct frequency.
    \figlabel{ls-model}}
\end{figure}

\subsection{Observational Noise}
\label{extensions-observational-noise}
Perhaps the most important modification of the periodogram is to
construct a periodogram that correctly handles measurement error in the data.
This involves changing the expression for the $\chi^2$ statistic in a manner
analogous to the change for a regression problem.
If there are Gaussian errors $\sigma_n$ on each observed $y_n$, we can re-write \eq{chi2-simple} in the following (standard) way:
\begin{equation}
  \chi^2(f) \equiv \sum_n \left(\frac{y_n - y_{model}(t_n;f)}{\sigma_n}\right)^2
  \eqlabel{chi2-with-errors}
\end{equation}
The periodogram constructed from this $\chi^2$ definition will more accurately
reflect the spectral power of noisy observations.
The effect of this modification on \eq{lomb-scargle-periodogram} is the addition
of a multiplicative weight $1/\sigma_n$ within each of the summations.
Early versions of this sort of modification appeared in \citet{Gilliland87}
and \citet{Irwin89}; \citet{Zechmeister09} showed that such a modification
does not change the desirable statistical properties of the resulting
periodogram.

@@

Another commonly-applied modification of the periodogram has variously been
called the {\it Date-compensated Discrete Fourier Transform} \citep{Ferraz-Mello81}, the {\it floating-mean periodogram} \citep{Cumming99}, and the {\it generalized Lomb-Scargle Method} \citep{Zechmeister09}, and involves adding an offset term to the sinusoidal model at each frequency:
\begin{equation}
  y_{model}(t;f) = y_{0, f} + A_f \sin(2 \pi f (t - \phi_f))
  \eqlabel{floating-mean-model}
\end{equation}
This is important, because the standard Lomb-Scargle approach assumes that the
data are pre-centered around the mean value of the signal.
In the past this was primarily accomplished by pre-centering the data before
the Lomb-Scargle fit using the mean estimated from the data points themselves.

With full phase coverage of the observed signal, this approach is generally
robust. However, due to selection effects and survey cadence, full phase
coverage can not always be guaranteed, which can potentially lead to situations
where the periodogram entirely misses peaks of interest.
\Fig{standard-vs-floatingmean} shows a simulated example of this: the data
consist of noisy observations of a sinusoidal signal in which the faintest
observations are omitted from the dataset (due to, {e.g.} a detection threshold).
Applying the standard Lomb-Scargle periodogram to pre-centered data leads to a
periodogram that misses the true period of 0.3 days (lower right panel).
Using the floating-mean model of \eq{floating-mean-model} yields a periodogram
that detects this true period (upper right panel).

@@

\citet{Zechmeister09} provides a detailed study of this
generalized/floating-mean periodogram, and shows that the beneficial
statistical properties of the standard periodogram apply here as well;
namely, periodogram power is independent and distributed according to a
simple, known function.

\subsection{Higher-order Fourier Models}

A further generalization involves multi-term Fourier models at each frequency:
rather than fitting just a single sinusoid at each frequency, we might fit a
partial Fourier series at each frequency:
\begin{equation}
  y_{model}(t;f) = y^{0}_f + \sum_{k=1}^K A_f^{(k)} \sin(2\pi k f (t - \phi_f^{(k)})).
\end{equation}
This type of multi-term Lomb-Scargle periodogram was studied in detail by
\citet{Bretthorst88}.

This kind of multiterm generalization is quite tempting, because it means that
the periodogram can fit models that are more complicated than

\todo{Show performance on eclipsing data}

\begin{figure}[ht]
  \centering
  \includegraphics[width=\textwidth]{fig21_kepler_multiterm}
  \caption{Multi-term models (left) and their corresponding periodograms (right)
    for the kepler data shown in \fig{kepler-data}.
    \figlabel{kepler-multiterm}}
\end{figure}

\begin{figure}[ht]
  \centering
  \includegraphics[width=\textwidth]{fig22_LINEAR_multiterm}
  \caption{Multi-term models (left) and their corresponding periodograms (right)
    for the LINEAR data shown in \fig{LINEAR-data}
    \figlabel{LINEAR-multiterm}}
\end{figure}

@@

\todo{Reference fast chi-square approach}
